---
title: "Data quality"
subtitle: "Coverage, outliers, and reconciliation checks across the warehouse"
---

```{r setup}
#| include: false
library(targets)
library(dplyr)
library(tidyr)
library(plotly)
library(reactable)
library(htmltools)
tar_source("R")

facts <- tryCatch(tar_read(facts_latest), error = function(e) tibble::tibble())
revs  <- tryCatch(tar_read(revisions),    error = function(e) tibble::tibble())
overrides <- tryCatch(
  sbm_load_overrides("data/manual_overrides/overrides.csv"),
  error = function(e) tibble::tibble()
)
```

::: {.callout-note}
This page runs the same battery of sanity checks that the
maintenance script `scripts/audit_data.R` produces on the command
line, but live against the current warehouse. Use it to spot
extraction regressions when new years roll in.
:::

## Coverage --- variables × jurisdictions

How many fiscal years of data we have per variable per jurisdiction.
Darker cells = more years.

```{r coverage}
if (nrow(facts) == 0L) {
  htmltools::p("No data.")
} else {
  cov <- facts |>
    count(variable_id, jurisdiction, name = "n_years") |>
    arrange(variable_id)

  plot_ly(cov,
          x = ~jurisdiction, y = ~variable_id, z = ~n_years,
          type = "heatmap",
          colorscale = list(c(0, "#f8f9fa"), c(0.2, "#bdd7e7"),
                            c(0.6, "#6baed6"), c(1, "#08519c")),
          hovertemplate = "<b>%{y}</b><br>%{x}: %{z} years<extra></extra>") |>
    plotly::layout(
      title = "Years of data per (variable, jurisdiction)",
      font  = list(family = "Inter, system-ui, sans-serif", size = 11L),
      xaxis = list(title = ""),
      yaxis = list(title = "", categoryorder = "category descending"),
      margin = list(l = 200L, r = 20L, t = 50L, b = 60L),
      height = 900
    )
}
```

## Magnitude check (FY 2024-25)

NSW=100 reference. If a state's headline ratios drift far from its
population/GSP share, that's a likely extraction error.

```{r magnitudes}
if (nrow(facts) == 0L) {
  htmltools::p("No data.")
} else {
  key_vars <- c("rev_total", "exp_total", "net_op_balance", "net_debt", "capex_pnfa")
  mag <- facts |>
    filter(variable_id %in% key_vars, fiscal_year == "2024-25") |>
    select(jurisdiction, variable_id, value_aud_mil) |>
    pivot_wider(names_from = variable_id, values_from = value_aud_mil) |>
    arrange(desc(rev_total))
  nsw_rev <- mag$rev_total[mag$jurisdiction == "NSW"]
  if (length(nsw_rev) > 0L && !is.na(nsw_rev)) {
    mag <- mag |> mutate(rev_pct_of_NSW = round(100 * rev_total / nsw_rev, 1))
  }
  mag |>
    mutate(across(where(is.numeric), \(x) round(x))) |>
    reactable(striped = TRUE, defaultPageSize = 8,
              defaultColDef = colDef(format = colFormat(separators = TRUE)))
}
```

## Sign-convention violations

Flags rows that violate basic expectations (e.g. negative revenue
or unemployment rate >25%).

```{r sign-checks}
sign_checks <- list(
  list(var = "net_debt",         rule = function(v) v < -50000, desc = "net_debt < -$50B (implausibly large net assets)"),
  list(var = "exp_total",        rule = function(v) v < 0,      desc = "exp_total < 0"),
  list(var = "rev_total",        rule = function(v) v < 0,      desc = "rev_total < 0"),
  list(var = "exp_employee",     rule = function(v) v < 0,      desc = "exp_employee < 0"),
  list(var = "capex_pnfa",       rule = function(v) v < 0,      desc = "capex_pnfa < 0"),
  list(var = "econ_unemployment",rule = function(v) v < 0 | v > 25, desc = "unemployment rate outside [0, 25]%")
)

flags <- purrr::map_dfr(sign_checks, function(s) {
  if (nrow(facts) == 0L) return(NULL)
  facts |>
    filter(variable_id == s$var, s$rule(value_aud_mil)) |>
    transmute(rule = s$desc, jurisdiction, fiscal_year, value_aud_mil, document_id)
})

if (nrow(flags) == 0L) {
  htmltools::p(htmltools::strong("Pass."), " No sign-convention violations across the warehouse.")
} else {
  reactable(flags, striped = TRUE, defaultPageSize = 15,
            columns = list(
              value_aud_mil = colDef(format = colFormat(separators = TRUE))
            ))
}
```

## Revenue subtotal reconciliation

`rev_total` should approximately equal the sum of its components
(taxation, GST, grants tied/untied, sales, interest, dividends,
other). Rows with mismatch > 20% are flagged. Common causes:

- LLM bundled tied + untied grants under one variable
- LLM extracted a different "Other revenue" line than what
  `rev_total` includes

```{r revtotal-reconciliation}
rev_components <- c("rev_taxation", "rev_gst", "rev_grants_tied", "rev_grants_untied",
                    "rev_sales_goods_svcs", "rev_interest", "rev_dividend", "rev_other")
if (nrow(facts) == 0L) {
  htmltools::p("No data.")
} else {
  rev_sum <- facts |>
    filter(variable_id %in% rev_components) |>
    group_by(jurisdiction, fiscal_year) |>
    summarise(rev_sum_components = sum(value_aud_mil, na.rm = TRUE),
              n_components = n(), .groups = "drop")
  rev_check <- facts |>
    filter(variable_id == "rev_total") |>
    select(jurisdiction, fiscal_year, rev_total = value_aud_mil) |>
    inner_join(rev_sum, by = c("jurisdiction", "fiscal_year")) |>
    mutate(diff_pct = round(100 * (rev_sum_components - rev_total) / rev_total, 1),
           flagged  = abs(diff_pct) > 20) |>
    filter(n_components >= 4) |>
    arrange(desc(abs(diff_pct)))
  reactable(rev_check, striped = TRUE, defaultPageSize = 10,
            filterable = TRUE,
            columns = list(
              rev_total = colDef(format = colFormat(separators = TRUE)),
              rev_sum_components = colDef(format = colFormat(separators = TRUE)),
              diff_pct = colDef(format = colFormat(suffix = "%")),
              flagged  = colDef(cell = function(v) if (v) "FLAG" else "")
            ),
            rowStyle = function(idx) {
              if (isTRUE(rev_check$flagged[idx])) list(background = "#fff4e6") else NULL
            })
}
```

## Year-over-year jumps

For headline variables, flags YoY changes greater than 50% (excluding
COVID-impacted years 2019-20 to 2021-22). Most flags are real, but
extreme jumps are worth verifying against the source.

```{r yoy}
if (nrow(facts) == 0L) {
  htmltools::p("No data.")
} else {
  key_vars <- c("rev_total", "exp_total", "net_op_balance", "net_debt", "capex_pnfa")
  yoy <- facts |>
    filter(variable_id %in% key_vars) |>
    group_by(jurisdiction, variable_id) |>
    arrange(fiscal_year, .by_group = TRUE) |>
    mutate(prev = lag(value_aud_mil),
           yoy_pct = round(100 * (value_aud_mil - prev) / abs(prev), 1)) |>
    ungroup() |>
    filter(!is.na(yoy_pct), abs(yoy_pct) > 50, abs(value_aud_mil) > 100,
           !(fiscal_year %in% c("2019-20", "2020-21", "2021-22"))) |>
    arrange(desc(abs(yoy_pct))) |>
    select(jurisdiction, variable_id, fiscal_year, prev, value_aud_mil, yoy_pct)
  if (nrow(yoy) == 0L) {
    htmltools::p(htmltools::strong("Pass."), " No YoY jumps > 50%.")
  } else {
    reactable(yoy, striped = TRUE, defaultPageSize = 15,
              columns = list(
                prev = colDef(format = colFormat(separators = TRUE)),
                value_aud_mil = colDef(format = colFormat(separators = TRUE)),
                yoy_pct = colDef(format = colFormat(suffix = "%"))
              ))
  }
}
```

## Unit / scale outliers

Rows where the value is more than 100× the median for the same
`(jurisdiction, variable)` over time --- usually means the LLM
picked the wrong row (e.g. a sub-line) instead of the headline.

```{r scale}
if (nrow(facts) == 0L) {
  htmltools::p("No data.")
} else {
  scale_check <- facts |>
    group_by(jurisdiction, variable_id) |>
    filter(n() >= 4, abs(value_aud_mil) > 0) |>
    mutate(med = median(abs(value_aud_mil), na.rm = TRUE),
           ratio = round(abs(value_aud_mil) / med, 4)) |>
    ungroup() |>
    filter(ratio > 100 | ratio < 0.01) |>
    arrange(desc(abs(log10(ratio)))) |>
    select(jurisdiction, variable_id, fiscal_year, value_aud_mil, med, ratio, document_id)
  if (nrow(scale_check) == 0L) {
    htmltools::p(htmltools::strong("Pass."), " No values >100x off their per-state median.")
  } else {
    reactable(scale_check, striped = TRUE, defaultPageSize = 15,
              columns = list(
                value_aud_mil = colDef(format = colFormat(separators = TRUE)),
                med = colDef(format = colFormat(separators = TRUE))
              ))
  }
}
```

## Manual overrides currently applied

Rows in `data/manual_overrides/overrides.csv` that exclude
known-bad LLM extractions from the warehouse.

```{r overrides}
if (nrow(overrides) == 0L) {
  htmltools::p("No overrides currently applied.")
} else {
  overrides |>
    transmute(Jurisdiction = jurisdiction,
              Variable     = variable_id,
              `FY`         = fiscal_year,
              Source       = document_id,
              `Override value` = value_aud_mil,
              Note         = notes) |>
    reactable(striped = TRUE, defaultPageSize = 10)
}
```

## Document FY drift

Sanity-check that each fact's fiscal year is within ±5 years of the
document's own fiscal year. Larger drift suggests a fiscal_year
mis-extraction.

```{r fy-drift}
if (nrow(facts) == 0L) {
  htmltools::p("No data.")
} else {
  drift <- facts |>
    mutate(doc_fy_start = as.integer(substr(stringr::str_extract(document_id, "[0-9]{4}-[0-9]{2}"), 1L, 4L)),
           row_fy_start = as.integer(substr(fiscal_year, 1L, 4L)),
           drift_years  = row_fy_start - doc_fy_start) |>
    count(drift_years, name = "n_rows") |>
    arrange(drift_years)
  plot_ly(drift, x = ~drift_years, y = ~n_rows,
          type = "bar",
          marker = list(color = "#1f3a93"),
          hovertemplate = "drift %{x} years: %{y:,} rows<extra></extra>") |>
    plotly::layout(
      title = "Distribution of (row FY) - (document FY)",
      font  = list(family = "Inter, system-ui, sans-serif", size = 12L),
      xaxis = list(title = "Drift in years"),
      yaxis = list(title = "Rows", tickformat = ","),
      margin = list(l = 60L, r = 20L, t = 50L, b = 60L)
    )
}
```
