---
title: "Australian State Budget Monitor"
subtitle: "Interactive fiscal positions across all eight states and territories"
---

```{r setup}
#| include: false
library(targets)
library(dplyr)
library(plotly)
library(reactable)
library(htmltools)
tar_source("R")

facts        <- tryCatch(tar_read(facts_latest),     error = function(e) tibble::tibble())
agg          <- tryCatch(tar_read(facts_aggregate),  error = function(e) tibble::tibble())
registry     <- tryCatch(tar_read(registry),         error = function(e) tibble::tibble())
downloaded   <- tryCatch(tar_read(downloaded_registry), error = function(e) tibble::tibble())
recon        <- tryCatch(tar_read(gfs_reconciliation), error = function(e) tibble::tibble())
```

## What this is

A reproducible R pipeline that extracts forward estimates and
historical actuals from every Australian state and territory Budget
Paper, MYEFO, and Final Budget Outcome, harmonises them against ABS
Government Finance Statistics, and renders them as this interactive
dashboard. Source code is at
[github.com/DavidAStephan/StateBudgetMonitor](https://github.com/DavidAStephan/StateBudgetMonitor);
the pipeline runs weekly and republishes the site automatically.

Drill in via the navigation above:

* **States** --- per-jurisdiction view of revenue, expenses, capex,
  and net debt with normalisation toggles and forward-estimate
  vintage trails.
* **Comparison** --- pick a variable, choose jurisdictions, switch
  between absolute / per-capita / share-of-GSP views.
* **Aggregate** --- sum across all eight jurisdictions.
* **Revisions** --- watch the forward estimate for any
  (jurisdiction, variable, year) triple move across successive
  Budget updates.
* **Wage policy** --- public-sector wage-policy decisions per state.
* **Methodology** --- sources, extraction approach, harmonisation,
  vintage discipline, ABS reconciliation, and known limitations.

## Pipeline coverage

```{r coverage}
coverage_cards <- function(facts, registry, downloaded) {
  htmltools::div(
    style = "display:flex; gap:1.5rem; margin:1rem 0;",
    htmltools::div(
      style = "flex:1; padding:1rem; background:#f8f9fa; border-radius:4px;",
      htmltools::strong("Documents tracked"),
      htmltools::tags$br(),
      htmltools::tags$span(style = "font-size:2rem;", nrow(registry))
    ),
    htmltools::div(
      style = "flex:1; padding:1rem; background:#f8f9fa; border-radius:4px;",
      htmltools::strong("Source URLs confirmed"),
      htmltools::tags$br(),
      htmltools::tags$span(
        style = "font-size:2rem;",
        if (nrow(registry) == 0L) "-"
        else sum(!is.na(registry$source_url))
      )
    ),
    htmltools::div(
      style = "flex:1; padding:1rem; background:#f8f9fa; border-radius:4px;",
      htmltools::strong("Facts in warehouse"),
      htmltools::tags$br(),
      htmltools::tags$span(style = "font-size:2rem;", nrow(facts))
    ),
    htmltools::div(
      style = "flex:1; padding:1rem; background:#f8f9fa; border-radius:4px;",
      htmltools::strong("Last as-of"),
      htmltools::tags$br(),
      htmltools::tags$span(
        style = "font-size:1.2rem;",
        format(Sys.Date(), "%d %b %Y")
      )
    )
  )
}

coverage_cards(facts, registry, downloaded)
```

## Snapshot: aggregate net debt

```{r snapshot}
if (nrow(agg) == 0L) {
  htmltools::p("No aggregate data yet --- add committed CSVs under `data/extracted/`.")
} else {
  agg |>
    filter(variable_id == "net_debt") |>
    arrange(fiscal_year) |>
    plot_ly(x = ~fiscal_year, y = ~value_aud_mil,
            color = ~is_forward_estimate,
            colors = c(`FALSE` = "#1f3a93", `TRUE` = "#5a6c7d"),
            type = "scatter", mode = "lines+markers",
            line = list(width = 3)) |>
    sbm_theme_fiscal(
      title       = "All states & territories --- net debt",
      yaxis_title = "AUD millions"
    )
}
```

## Per-jurisdiction coverage

```{r coverage-table}
if (nrow(facts) == 0L) {
  htmltools::p("No facts yet.")
} else {
  facts |>
    count(jurisdiction, name = "facts") |>
    arrange(desc(facts)) |>
    reactable(
      striped         = TRUE,
      defaultPageSize = 10
    )
}
```
