---
title: "Methodology"
subtitle: "Sources, extraction, harmonisation, vintage discipline, and known limitations"
---

```{r setup}
#| include: false
library(targets)
library(dplyr)
library(reactable)
library(htmltools)
tar_source("R")

recon      <- tryCatch(tar_read(gfs_reconciliation), error = function(e) tibble::tibble())
registry   <- tryCatch(tar_read(registry),          error = function(e) tibble::tibble())
dictionary <- tryCatch(tar_read(dictionary),        error = function(e) tibble::tibble())
cfg        <- tryCatch(tar_read(cfg),               error = function(e) list())
tolerance_pct <- if (length(cfg) > 0L) {
  cfg$validation$gfs_tolerance_pct %||% 2
} else {
  2
}
```

## What this project does

A public, reproducible R pipeline that:

1. Maintains a registry of every Australian state and territory Budget
   Paper, MYEFO, and Final Budget Outcome from FY2014-15 onwards.
2. Downloads each document and extracts the General Government Sector
   forward-estimates table.
3. Harmonises line items across jurisdictions against the ABS
   Government Finance Statistics chart of accounts.
4. Stores every fact in a DuckDB warehouse keyed on its source
   document, so that any past "as-at" view of the data can be
   reconstructed (vintage discipline).
5. Renders this dashboard with Quarto, fully static.

## Sources

* **Primary --- state Treasury / Budget websites.**
  Budget Papers, MYEFOs, and Final Budget Outcomes for each
  jurisdiction. URLs change over time; the canonical list lives in
  [`inst/document_registry.csv`](https://github.com/DavidAStephan/StateBudgetMonitor/blob/main/inst/document_registry.csv).
* **Cross-validation --- ABS Government Finance Statistics.**
  Annual (cat. 5512.0) and quarterly (cat. 5519.0.55.001), pulled
  via `readabs`. Where ABS has migrated a catalogue to data cubes,
  we fall back to `readabs::download_abs_data_cube()`.
* **Normalisation --- ABS population (cat. 3101.0) and ABS state
  accounts (cat. 5220.0).** Powers the per-capita and share-of-GSP
  views.

## Extraction

* **Rule-based first** via `tabulapdf`, per-jurisdiction parsers
  under `R/extract/extract_<state>.R`. Each parser is version-tagged;
  the `parser_version` is recorded against every fact in the
  warehouse.
* **LLM-assisted backfill** via `ellmer` (Anthropic API) for
  documents where rule-based parsing fails or for unusual layouts.
  **The LLM-assisted step runs only on a developer machine.** Its
  structured outputs are committed to `data/extracted/`, and the
  production CI pipeline reads from those CSVs --- CI never calls an
  LLM.
* **Manual overrides** for edge cases live in
  `data/manual_overrides/overrides.csv` and take precedence in the
  warehouse build.

The CSV contract is documented in
[`data/extracted/README.md`](https://github.com/DavidAStephan/StateBudgetMonitor/blob/main/data/extracted/README.md).

## Harmonisation

State charts of accounts differ; the same economic concept may be
called different things in different jurisdictions. The canonical
dictionary is a long-format mapping with one row per
`(canonical_variable_id, jurisdiction)`, pairing each canonical
variable to the state's source line-item text plus mapping notes.

For revenue, expense, and capex, the canonical reference is ABS GFS
since that's what flows into national accounts and is what the RBA
ultimately cares about.

```{r dict-summary}
if (nrow(dictionary) == 0L) {
  htmltools::p("Dictionary not loaded.")
} else {
  dictionary |>
    count(category, sub_category, name = "n_jurisdictions") |>
    arrange(category, sub_category) |>
    reactable(
      defaultPageSize = 15,
      searchable      = TRUE,
      striped         = TRUE,
      defaultColDef   = colDef(minWidth = 80)
    )
}
```

## Vintage discipline

Every fact in the warehouse is keyed on the document it was
extracted from. That document's release date lives in
`dim_documents`. To ask "what was knowable about NSW net debt for
FY2025-26 as at 1 March 2024?", filter facts to those whose document
had `release_date <= 2024-03-01`, then take the most recent per
`(jurisdiction, variable_id, fiscal_year)`. This is the same pattern
used by the sister project
[NOM_Nowcast](https://github.com/DavidAStephan/NOM_Nowcast).

The revisions page exposes this directly: pick a variable and a
forecast year, see how each Budget update has moved the estimate.

## ABS GFS reconciliation

For headline variables (revenue, expenses, balance, net debt) we
expect our extracted figures to track ABS GFS within
`r format(round(tolerance_pct, 1), nsmall = 1)`%.
Discrepancies above tolerance are flagged here.

```{r recon-table}
if (nrow(recon) == 0L) {
  htmltools::div(
    htmltools::p("No reconciliation rows yet."),
    htmltools::p(htmltools::tags$em(
      "ABS GFS catalogue 5512.0 has migrated to data cubes. The fetcher ",
      "is wired up via readabs::download_abs_data_cube() but currently ",
      "returns no rows --- see STATUS.md for the catalogue/cube args that ",
      "need ABS-current verification. Reconciliation populates here once ",
      "the upstream fetch succeeds."
    ))
  )
} else {
  recon |>
    arrange(desc(abs(diff_pct))) |>
    reactable(
      defaultPageSize = 15,
      searchable      = TRUE,
      filterable      = TRUE,
      striped         = TRUE,
      rowStyle        = function(idx) {
        if (isTRUE(recon$flagged[idx])) list(background = "#fff4e6") else NULL
      }
    )
}
```

## Document registry coverage

```{r registry-coverage}
if (nrow(registry) == 0L) {
  htmltools::p("Registry not loaded.")
} else {
  registry |>
    count(jurisdiction, doc_type, name = "n_documents") |>
    tidyr::pivot_wider(names_from = doc_type, values_from = n_documents,
                       values_fill = 0) |>
    arrange(jurisdiction) |>
    reactable(
      striped = TRUE,
      defaultPageSize = 8
    )
}
```

## Known limitations

* **PDF layout changes are the highest-risk failure mode.** Every
  parser is version-tagged; cross-validation against ABS GFS surfaces
  silent regressions.
* **Per-capita / share-of-GSP normalisations use ABS annual
  estimates.** Quarterly population estimates are available; we use
  the end-of-FY (30 June) ERP value.
* **Wage policy text extraction is best-effort.** The authoritative
  source is each Budget Paper itself; values shown here are seed /
  preliminary until each is verified against BP1.
* **Registry URLs drift over time.** Most rows are seeded with
  landing-page URLs; specific document URLs are populated as each is
  parsed. The downloader skips rows without a URL rather than
  blocking.

## Reproducibility

The project is an installable R package (`statebudgetmonitor`) with a
locked `renv.lock`. To run end-to-end:

```r
renv::restore()
targets::tar_make()
quarto::quarto_render()
```

See the project
[README](https://github.com/DavidAStephan/StateBudgetMonitor#readme)
for fork-and-deploy instructions.
